%%
% 说明
% modifier: 郑齐(Tidyzq)
% update date: 2018-4-30

\begin{center}
  {\zihao{2}\heiti\bfseries 说\hspace{2\ccwd}明}
\end{center}

\fontsize{15}{20}\selectfont

\zihao{4}\songti
\begin{enumerate}[label=\arabic*.,leftmargin=*]
  \item 毕业论文（设计）的写作格式要求请参照《中山大学本科生毕业论文的有关规定》和《中山大学本科生毕业论文（设计）写作与印制规范》。
  \item 除完成毕业论文（设计）外，还须填写三份表格：
  \begin{enumerate}[align=left,leftmargin=!,labelsep=0mm,label=（\arabic*）]
    \item 表一\hspace{\ccwd}毕业论文（设计）开题报告；
    \item 表二\hspace{\ccwd}毕业论文（设计）过程检查情况记录表；
    \item 表三\hspace{\ccwd}毕业论文（设计）答辩情况登记表。
  \end{enumerate}
  \item 上述表格均可从教务部主页的“下载中心”处下载，如表格篇幅不够，可另附纸。每份毕业论文（设计）定稿装订时应随同附上这三份表格。
  \item 封三是毕业论文（设计）成绩评定的主要依据，请认真填写。
\end{enumerate}
\vspace{1\baselineskip}

\begin{center}
  {\zihao{2}\bfseries Instruction}
\end{center}

\zihao{4}
\begin{enumerate}[label=\arabic*.,leftmargin=*]
  \item Please refer to ‘\textit{The Guidelines to Undergraduate Graduation Thesis (Design) at Sun Yat-sen University}’and ‘\textit{The Writing and Printing Format of Undergraduate Graduation Thesis(Design) at Sun Yat-sen University}’ for anything about the thesis format.
  \item Three forms should be filled up before the submission of the thesis (design)：
  \begin{enumerate}[align=left,leftmargin=!,labelsep=0mm,label=（\arabic*）]
    \item Form 1: Research Proposal of Graduation Thesis.
    \item Form 2: Process Check-up Form.
    \item Form 3: Thesis Defense Performance Form.
  \end{enumerate}
  \item All the above forms could be downloaded on the website of the Office of Education Administration. If there is not enough space in the form, please add extra sheets. Each thesis (design) should be submitted together with the three forms.
  \item The form on the inside back cover is the grading sheet. Please fill it up before submission.
\end{enumerate}
